---
title: "T4: Mode Choice"
format: 
  html:
    theme: minty
    toc: true
    toc-location: left
editor: visual
---

The purpose of this assignment is for you to

-   Run the mode choice step of TDM23 for both the existing (2019) condition and an alternative you've developed (it could be the same one you used for previous assignments, or a new one), and

-   Identify changes in mode choice that result from your alternative scenario.

## Load libraries

This analysis uses the following packages:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(DBI)
library(here)
library(leaflet)
library(maptiles)
library(tidyterra)
library(knitr)
library(tigris)
library(omxr)
library(scales)
library(RColorBrewer)

sf::sf_use_s2(FALSE)
```

## Identify Somerville zones

lorem ipsum

```{r}

somerville_zones <- here("model",
             "inputs",
             "zonal",
             "shp",
             "CTPS_TDM23_TAZ_2017g_v202303.shp") |>
  st_read(quiet = TRUE) |>
  st_drop_geometry() |>
  select(taz_id, town) |>
  filter(town == "SOMERVILLE")

```

## Load trip matrices in R

You can can read the files you just exported into R. Here is how you can load the trip matrix for the base scenario and view the first few rows.

```{r, message=FALSE}

base_mc <- here("model",
                "outputs",
                "Base",
                "_demand",
                "mc",
                "hh_trips_daily.omx") |>
  read_all_omx()

somerville_mc_base <- base_mc |>
  filter(origin %in% somerville_zones$taz_id |
           destination %in% somerville_zones$taz_id)

head(base_mc) |>
  kable(digits = 1)

```

We've imported it as a dataframe with the following columns:

-   **origin:** The TAZ ID of the trip productions (they're using origin/destination terms here, but these are still productions/attractions at this point in the model workflow).

-   **destination:** The TAZ ID of the trip attractions

-   **bk:** The number of daily trips by bike.

-   **da:** The number of daily trips by driving alone.

-   **rs:** The number of trips by ride-sourcing (e.g. Uber or Lyft).

-   **s2:** The number of private car trips with two people per vehicle.

-   **s3:** The number of private car trips with three or more people per vehicle.

-   **sb:** The number of bike share trips.

-   **ta:** The number of transit trips with car as an access or egress mode (e.g. park-and-ride trips)

-   **tw:** The number of transit trips with walking as the access/egress mode.

-   **wk:** The number of walk trips.

And we can also load the production-attraction matrix for the alternative scenario.

```{r, message=FALSE}

alt_mc <- here("model",
             "outputs",
             "UBI-1",
             "_demand",
             "mc",
             "hh_trips_daily.omx") |>
  read_all_omx()

head(alt_mc) |>
  kable(digits = 1)

somerville_mc_alt <- alt_mc |>
  filter(origin %in% somerville_zones$taz_id |
           destination %in% somerville_zones$taz_id)

# Columns to sum
cols_to_sum <- c("da", "rs", "s2", "s3", "sb", "ta", "tw", "wk", "bk")

# --- 1. Create total summary dataframes ---

# ALT totals
somerville_mc_alt_totals <- as.data.frame(t(sapply(cols_to_sum, function(col) {
  sum(somerville_mc_alt[[col]], na.rm = TRUE)
})))
colnames(somerville_mc_alt_totals) <- paste0(cols_to_sum, "_total")

# BASE totals
somerville_mc_base_totals <- as.data.frame(t(sapply(cols_to_sum, function(col) {
  sum(somerville_mc_base[[col]], na.rm = TRUE)
})))
colnames(somerville_mc_base_totals) <- paste0(cols_to_sum, "_total")

# --- 2. Rename columns to track source ---

colnames(somerville_mc_alt_totals) <- paste0(colnames(somerville_mc_alt_totals), "_alt")
colnames(somerville_mc_base_totals) <- paste0(colnames(somerville_mc_base_totals), "_base")

# --- 3. Combine into comparison dataframe ---

somerville_mc_comparison <- cbind(somerville_mc_alt_totals, somerville_mc_base_totals)

# --- 4. Compute differences ---

# Get base variable names
base_names <- gsub("_total_base$", "", names(somerville_mc_comparison)[grepl("_total_base$", names(somerville_mc_comparison))])

# Compute difference for each variable (corrected)
somerville_mc_difference <- as.data.frame(t(sapply(base_names, function(name) {
  somerville_mc_comparison[[paste0(name, "_total_alt")]] - somerville_mc_comparison[[paste0(name, "_total_base")]]
})))

# Name the columns
colnames(somerville_mc_difference) <- paste0(base_names, "_difference")



```

## Calculate Mode Shares

```{r}

mode_share_by_zone <- function(mc_table) {
  mc_by_p_zone <- mc_table |>
    group_by(origin) |>
    summarise(across(everything(), list(p = sum))) |>
    rename(zone = origin) |>
    select(zone, bk_p, da_p, rs_p, s2_p, s3_p, sb_p, ta_p, tw_p, wk_p)
              
  mc_by_a_zone <- mc_table |>
    group_by(destination) |>
    summarise(across(everything(), list(a = sum))) |>
    rename(zone = destination) |>
    select(zone, bk_a, da_a, rs_a, s2_a, s3_a, sb_a, ta_a, tw_a, wk_a)
  
  full_join(mc_by_p_zone, mc_by_a_zone) |>
    mutate(bk = bk_p + bk_a,
           da = da_p + da_a,
           rs = rs_p + rs_a,
           s2 = s2_p + s2_a,
           s3 = s3_p + s3_a,
           sb = sb_p + sb_a,
           ta = ta_p + ta_a,
           tw = tw_p + tw_a,
           wk = wk_p + wk_a) |>
    mutate(total = bk + da + rs + s2 + s3 + sb + ta + tw + wk) |>
    mutate(bk_pct = bk / total,
           da_pct = da / total,
           rs_pct = rs / total,
           s2_pct = s2 / total,
           s3_pct = s3 / total,
           sb_pct = sb / total,
           ta_pct = ta / total,
           tw_pct = tw / total,
           wk_pct = wk / total) |>
    select(zone,
           total,
           bk,
           da,
           rs,
           s2,
           s3,
           sb,
           ta,
           tw,
           wk)
}
```

```{r, message=FALSE}

base_mc_by_zone_total <- mode_share_by_zone(somerville_mc_base)

alt_mc_by_zone_total <- mode_share_by_zone(somerville_mc_alt)
```

## Check for differences

```{r, message=FALSE}
base_mc_by_zone_total <- base_mc_by_zone_total |>
  rename_with(~ paste0("base_", .), -all_of("zone"))

alt_mc_by_zone_total <- alt_mc_by_zone_total |>
  rename_with(~ paste0("alt_", .), -all_of("zone"))

mc_comparison <- full_join(base_mc_by_zone_total, alt_mc_by_zone_total) |>
  replace_na(list(base_total = 0,
                  alt_total = 0)) |>
  filter(alt_total + base_total > 0) |>
  mutate(diff_bk = alt_bk - base_bk,
         diff_da = alt_da - base_da,
         diff_rs = alt_rs - base_rs,
         diff_s2 = alt_s2 - base_s2,
         diff_s3 = alt_s3 - base_s3,
         diff_sb = alt_sb - base_sb,
         diff_ta = alt_ta - base_ta,
         diff_tw = alt_tw - base_tw,
         diff_wk = alt_wk - base_wk)

```

```{r, message=FALSE}
somerville_zones <- here("model",
             "inputs",
             "zonal",
             "shp",
             "CTPS_TDM23_TAZ_2017g_v202303.shp") |>
  st_read(quiet = TRUE) |>
  st_transform("WGS84") |>
  rename(zone = taz_id) |>
  left_join(mc_comparison) |>
  filter(abs(diff_da) > 0.001) |>
  filter(town == "SOMERVILLE")

base_map <- get_tiles(somerville_zones,
                      provider = "CartoDB.Positron",
                      zoom = 13,
                      crop = TRUE)

# Create and assign the plot
changes_da_somerville <- ggplot(somerville_zones) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(aes(fill = diff_da),
          color = NA,
          alpha = 0.7) +
  scale_fill_gradient2(name = "Difference in share\nof drive-alone trips") +
  theme_void()

# Save the plot
ggsave("changes_da_somerville.png", plot = changes_da_somerville, width = 8, height = 6, dpi = 300)

```

```{r}

mc_comparison |> 
  filter(diff_da < .5) |>
  select(base_total,
         alt_total,
         diff_da,
         diff_s2,
         diff_s3,
         diff_rs,
         diff_tw,
         diff_ta,
         diff_bk,
         diff_wk) |>
  t() |>
  kable(digits = 2)
```
